---
title: An R Markdown document converted from "course-notes/01-review.ipynb"
output: html_document
---

# Review for PSQF 6243

This serves as a non-exhaustive review for the course. These are elements that I assume you have knowledge of prior to starting the course. 

+ Variable vs constant attributes
+ Types of variables (ie., nominal, ordinal, integer, ratio)
+ Descriptive Statistics (eg., mean, median, standard deviation, variance, percentiles)
+ Higher order moments (eg., skewness and kurtosis)
+ Exploring/summarizing univariate distributions (eg., histogram or density figure)
+ What is a statistical model? Why do we use them?
+ Population vs Sample


## Examples

Mario Kart 64 world record data:

| variable         | class     | description                     |
|:-----------------|:----------|:--------------------------------|
| track            | character | Track name                      |
| type             | factor    | Single or three lap record      |
| shortcut         | factor    | Shortcut or non-shortcut record |
| player           | character | Player’s name                   |
| system_played    | character | Used system (NTSC or PAL)       |
| date             | date      | World record date               |
| time_period      | period    | Time as `hms` period            |
| time             | double    | Time in seconds                 |
| record_duration  | double    | Record duration in days         |

```{r}
# load some libraries
library(tidyverse)
library(ggformula)
library(lubridate)
library(mosaic)
library(e1071)

theme_set(theme_bw(base_size = 18))

# load in some data
mariokart <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-05-25/records.csv') %>%
    mutate(year = year(date),
           month = month(date),
           day = month(date))

head(mariokart)
```

```{r}
# univariate distribution of time
gf_histogram(~ time, data = mariokart, bins = 30) %>% 
   gf_labs(x = "Time (in seconds)")
```

```{r}
gf_density(~ time, data = mariokart) %>% 
   gf_labs(x = "Time (in seconds)")
```

```{r}
df_stats(~ time, data = mariokart, mean, median, sd, skewness, kurtosis, quantile(probs = c(0.1, 0.5, 0.9)))
```

# Bivariate Association

```{r}
cor(time ~ record_duration, data = mariokart)
```

```{r}
gf_point(time ~ record_duration, data = mariokart) %>%
  gf_labs(x = "How long the record was held",
          y = "Time (in seconds)")
```

## Questions
1. What is problematic about the analyses above? Why?
2. What could be done to improve the analyses above?

