---
title: An R Markdown document converted from "course-notes/04-least-squares-simulation.ipynb"
output: html_document
---

## Example to show least squares minimization

This little example is meant as a way to show the least square really minimizes the criterion, $ \sum \left( Y - \hat{Y} \right)^2 $.

In this example, we will generate some data so that we know what the truth is. Then, upon data generation, we will compute a bunch of different values for the linear slope and y-intercept. For each combination of the y-intercept and slope, I will compute the sum of squares error depicted above. 

```{r}
library(tidyverse)
library(simglm)

theme_set(theme_bw(base_size = 18))

set.seed(2021)

sim_arguments <- list(
    formula = y ~ x,
    fixed = list(x = list(var_type = 'continuous', mean = 100, sd = 20)),
    error = list(variance = 100),
    sample_size = 500,
    reg_weights = c(5, .5)
)

sim_data <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

head(sim_data)
```

```{r}
library(ggformula)

gf_point(y ~ x, data = sim_data, size = 4) %>%
  gf_smooth(method = 'lm')
```

```{r}
sim_lm <- lm (y ~ x, data = sim_data)
coef(sim_lm)
```

```{r}
y_intercept <- seq(0, 15, by = .25)
slope <- seq(0, 1.5, by = .05)

conditions <- rbind(expand.grid(y_intercept = y_intercept, 
                          slope = slope),
                          coef(sim_lm))

tail(conditions)
dim(conditions)
```

```{r}
gf_point(y ~ x, data = sim_data, size = 4) %>%
  gf_smooth(method = 'lm') %>%
  gf_abline(slope = ~slope, intercept = ~y_intercept, data = slice(conditions, 1), linetype = 2, size = 2) %>%
  gf_abline(slope = ~slope, intercept = ~y_intercept, data = slice(conditions, 855), linetype = 2, color = 'lightgreen', size = 2) %>%
  gf_refine(coord_cartesian(xlim = c(0, 160), ylim = c(0, 120)))

```

```{r}
sum_square_error <- function(conditions, sim_data) {
    fitted <- conditions[['y_intercept']] + conditions[['slope']] * sim_data[['x']]

    deviation <- sim_data[['y']] - fitted

    sqrt((sum(deviation^2) / (nrow(sim_data) - 2)))
}

sum_square_error(conditions[1892, ], sim_data)
summary(sim_lm)$sigma
```

```{r}
library(future)

plan(multicore)

conditions$sse <- unlist(lapply(1:nrow(conditions), function(xx) sum_square_error(conditions[xx, ], sim_data)))

head(conditions)
```

